---
title: "University of Oregon Gym Attendance: An Analysis of Student Behavior and Decision Making"
author: "Hunter Wright"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library(pacman)
p_load(tidyverse, forcats, kableExtra, stargazer, tseries, lmtest, sandwich, scales)

count <- read.csv("cleanData/count.csv")
members <- read.csv("cleanData/membership.csv")

```
```{r factor releveling, include=FALSE}
members <- members %>%
  mutate(day = fct_relevel(day, "Monday"))
count <- count %>%
  mutate(day = fct_relevel(day, "Monday")) %>%
  mutate(month = fct_relevel(month, "October")) %>%
  mutate(quarter = fct_relevel(quarter, "Fall"))

count$date <- as.Date(count$date)

count_open <- count %>%
  filter(closed == 0)
```

Notes:

- expand analysis/interpretation
- incorporate membership controls
- how to preview results?
- add formulas to represent behavioral economics concepts?
- how to expand literature review?
- formatting needs fix (last thing to do)


# Introduction

The health and fitness industry has seen substantial growth since COVID-19 pandemic restrictions lifted. According to the 2024 Health and Fitness Association Global Report, key developed markets around the world are seeing record high rates of gym memberships among their adult population. The United States leads with 23.7% of surveyed adults saying they belonged to a gym or health club. On top of this, markets with historically lower participation are seeing record growth as health and fitness enters the forefront of the public consciousness. However, having a gym membership and actually going to the gym are two different things.

The field of behavioral economics primarily focuses on how individuals make risky decisions with monetary payoffs. Should someone spend their paycheck or invest it? What does an individual expect to gain from buying a lottery ticket? Money, or capital is a scarce resource that agents must take into account when making decisions that affect their future. In this thesis, I look to explore how agents make decisions with another scarce resource: time. While there is a wide disparity in wealth, where decisions involving money are more important to some than others, everyone has the same 24 hours in a day, and must make risky decisions based on that constant. One of the main purposes taking classes at a university is to develop a students' education, skills, general knowledge, health and life experience. In economics, these are considered useful personal attributes and are known as **human capital**. College students look to develop their human capital taking classes, joining clubs, doing internships, and exercising, among many other things. All of these activities take time though, and a rational student must make important decisions on how to spend that time. On any given day of the week, a college student might have class, work, homework, and other responsibilities that take up most of their day. Coupled with basic tasks like eating and traveling, many students find their only free time on the weekends. 

According to the University of Oregon's Student Recreation Center (SRC) website, “The Student Recreation Center is consistently busy Monday–Thursday during the term… Friday–Sunday, the Rec has much less traffic.” Most students have busy schedules during the week, and much more time on the weekend, so why does SRC traffic decrease when individuals have more time? In this thesis, I look to analyze short and long-term trends in SRC attendance to discuss how students display behaviors consistent with the behavioral economics concepts of mental accounting, present bias, and loss aversion when choosing whether or not to go to the gym. 

After regressing gym attendance on variables like day of the week, season, and term, controlling for variables like previous day's attendance and membership type, I find statistically significant negative relationships across all time variables. This indicates that across time periods like weeks, months, and terms, attendance starts high and then decreases at varying rates as the time period elapses. These results reinforce the ideas that students at the University of Oregon are showing patterns of present bias and loss aversion in their intertemporal decision making with respect to gym attendance.


### Literature Review

The primary literature relating to this thesis is Prospect Theory: An Analysis of Decision Under Risk, by Kahneman and Tversky (1979). This paper serves as a critique of expected utility theory by providing evidence that humans do not act completely rationally when presented with decisions that involve risk. While Kahneman and Tversky were focused mainly on monetary decisions under risk, I focus on the scarcity of time as another form of currency, where students risk the development of their human capital based on how they utilize limited time. Other similar literature related to gym attendance focuses mainly on incentives. Incentives are another behavioral economic concept originating from Nudge Theory in the book Nudge: Improving Decisions About Health, Wealth, and Happiness by Thaler and Sunstein (2008).

In discussing present bias in the Concepts section, I reference "Doing It Now or Later" by O'Donoghue and Rabin (1999), where the concept was formally introduced. For loss aversion, I reference "Loss Aversion in Riskless Choice: A Reference-Dependent Model" by Tversky et. al (1991) and "The effect of myopia and loss aversion on risk taking: An experimental test" by Thaler et. al (1997).


# Concepts in Behavioral Economics Theory

Behavioral economics is the study of how people in the real world approach decisions in an irrational manner. Contrary to what standard model economics believes, people don't disregard emotion, habit, and other feelings when making decisions. Though behavioral economics primarily focuses on how individuals make decisions with limited money, this thesis shows that individuals may treat limited time in a similar manner.  

### The Standard Economic Model

The standard economic model predicts that all individuals are rational agents that maximize their utility subject to a constraint. In this case, the agent's constraint is the number of hours in a day, and with so few hours of free time in a weekday, we would not expect a rational agent to prioritize going to the gym. In other words, if this model reflected reality, we would observe a decrease in gym attendance during the week where the opportunity cost of an hour is higher, and increase in gym attendance on the weekend where the opportunity cost of an hour is cheaper. However, as Figure 1 shows, this tends not to be the case.

```{r average fall 2024 attendance plot, fig.height=3, fig.width=5, fig.align = 'center'}

count %>%
  mutate(year = year(date)) %>%
  filter(year == 2024,
         quarter == "Fall") %>%
  group_by(day) %>%
  summarize(average = mean(total)) %>%
  ungroup() %>%
  mutate(
    day = factor(day, levels = c("Monday","Tuesday","Wednesday",
                                "Thursday","Friday","Saturday","Sunday"))) %>%
  ggplot(aes(x = day, y = average, group = 1)) + 
  geom_line() +
  geom_point(size = 2) +
  scale_y_continuous(labels = comma_format()) +
  labs(
    x = NULL,
    y = "Attendance",
    title = "Figure 1: Average SRC Daily Attendance, Fall 2024"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 7),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 7)
  )

```

### Mental Accounting



### Present Bias
 
Individuals experiencing a dilemma with ramifications that unfold over time (known as intertemporal choice) may engage in temporal discounting, where they place disproportionately less value on future rewards, even if they are greater than present rewards. This is known as **present bias**, formally introduced by  O'Donoghue and Rabin in the 1999 paper "Doing It Now or Later." For example, someone may choose to use their disposable income to consume goods (like buying a car), instead of putting it in a savings account, even though investing will ultimately net them more utility in the future. In this thesis, I aim to interpret present bias in how individuals utilize their time. For many people, going to the gym is hard and its payoffs take time. If someone only has two days a week to relax, they might skip the gym on the weekend and spend their time relaxing, because the immediate gratification of relaxing is worth more to them than the future benefit of a consistent gym routine. This is a significant departure from the standard economic model's prediction, which interprets this behavior as irrational.

### Loss Aversion

Loss aversion stems from an agent's perceived utility from losses and gains measured relative to a reference point, where a reference point is a baseline or standard of comparison. In their 1991 paper, Tversky et. al state, "the function is steeper in the negative than in the positive domain; losses loom larger than corresponding gain." The function in question represents the utility of an action relative to a reference point. In other words, an agent will suffer a larger utility loss from losing `$`5 than they will gain from receiving `$`5. In the context of this thesis, a student may spend their Saturdays relaxing. The concept of this time being set aside for relaxation becomes the student's reference point, and the utility of anything they have to do that deviates from relaxation will be compared to the utility of relaxing. This concept is represented by the following formula:

$$ 
v(x) =
\begin{cases}
  x,                    & x \ge 0,\\
  -\lambda\,(-x),       & x < 0.
\end{cases}
$$

Here, $x$ represents the gain ($x>0$) or loss ($x < 0$) relative to the reference point. $\lambda$ represents the loss aversion coefficient, which acts as a multiplier for negative utility. For a student relaxing on the weekend, $x$ represents the utility difference between relaxing and going to the gym (or doing anything else that represents hard work). If $\lambda > 1$ then the student decides to relax instead of going to the gym. I find the opposite situation to be true during the week, where the reference point is classes, homework, and other responsibilities. During the week, relaxing becomes the loss and hard work like studying and the gym provide more utility. 

### Myopic Loss Aversion

Myopic loss aversion is the combination of mental accounting, present bias, and loss aversion. When an investor is too focused on short term losses, they may make decisions at the expense of long term gains, like panic selling a stock when it crashes in the short term. In their 1997 paper, "The effect of myopia and loss aversion on risk taking: An experimental test," Thaler et al. defined myopic loss aversion as "the combination of a greater sensitivity to losses than to gains and a tendency to evaluate outcomes frequently."

# Data

The University of Oregon’s Student Recreation Center is the primary facility on campus for students to engage in physical activities like weightlifting, swimming, basketball, and other indoor sports. Students, faculty, staff, and community members alike enjoy more than 36,000 square feet of strength training space, a 12-lane swimming pool, and nearly two dozen courts and gymnasiums, among many other amenities. During normal hours, the SRC is open from 6am - 11pm on weekdays and 9am-9pm on weekends. Almost every person that enters the SRC must scan their ID through one of the turnstiles. These machines then record the time and date of every person who walks through. Table 1 shows the summary statistics for daily entries between January 2015 and December 2024 for days where the SRC is open. This accounts for about 81% of the days in the data, which includes a period between March 18th, 2020 and September 17th, 2020 where the SRC closed due to COVID-19.

```{r count summary statistics table}

# closed days are removed
.summary_stats <- count_open %>%
  summarise(
    Mean = mean(total),
    Median = median(total),
    SD = sd(total),
    Min = min(total),
    Max = max(total)
  )

kable(.summary_stats, caption = "Summary Statistics") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(0, align = "c") %>%
  row_spec(1, align = "c")
```

Table 2 shows the average daily proportion of members. Students make up the vast majority of the SRC's daily entrants, which is why most of the analysis is targeted towards their behavior. It should be noted that "Off-Term Students" are simply students who are not regarded as full-time students because they are taking less than 12 credits of classes. This number is skewed by the heavy increase in this membership type that occurs every summer when students continue to use the SRC while on summer break.

```{r membership summary statistics table}

members_open <- members %>%
  filter(closed == 0)

.member_columns <- c("students", "student_offterm", "faculty_staff", "pe_staff", "alumni", "nonmembers", "community_members", "other")

.mem_sumstats <- members_open %>% 
  summarise(
    across(all_of(.member_columns),
           ~ mean(.x / total, na.rm = TRUE)) 
  ) %>%
  rename("Student" = students,
         "Faculty/Staff" = faculty_staff,
         "PE Staff" = pe_staff,
         "Alumni" = alumni,
         "Nonmember" = nonmembers,
         "Community Member" = community_members,
         "Off-Term Student" = student_offterm,
         "Other" = other) %>% 
  pivot_longer(everything(),
               names_to  = "Membership Type",
               values_to = "Average Daily Share") %>% 
  mutate(`Average Daily Share` = percent(`Average Daily Share`, accuracy = 0.1)) %>%
  arrange(desc(`Average Daily Share`))


# need to fix row separations
kable(.mem_sumstats, caption = "Average Daily Entrances by Member") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(0:8, align = "c")
```

Two data sets were collected and used for this analysis. The first displayed simple daily totals grouped by time of day, while the other displayed these totals grouped by membership type. The majority of the analysis, plots and tables were produced using the first data set. The second data set was used for its own analysis, which is labeled. 

There is a discrepancy in daily total entrances between the data sets. For some unknown reason, the membership data set tends to under count daily entrances, averaging about 277 fewer entrances a day, which is why it is only used for its particular analysis section. 


- talk about possible data collection issues.

# Methods

For my analysis, I maintain my use of the recorded daily entrances between January 2015 and December 2024, without

To analyze both short and long‑run attendance dynamics, I regress daily gym patron attendance on lagged attendance, weekday, week‑of‑term, month, quarter, post‑closure status, and finals/dead‑week indicators. 

$$
\hat{\text{total}}_{t} = \beta_0 + \beta_1\,\text{total}_{t-1} \;+\; \sum_{d=1}^{6}\gamma_d\,\text{Day}_{d,t} \;+\;\beta_2\,\text{Week}_t \;+\;
\sum_{m=2}^{12}\delta_m\,\text{Month}_{m,t} \;+\; \sum_{q=2}^{4}\theta_q\,\text{Quarter}_{q,t}\;+\; \beta_3\,\text{AfterClosed}_t
$$ 
$$
\;+\; \beta_4\,\text{FinalsWeek}_t \;+\; \beta_5\,\text{DeadWeek}_t \;+\; \varepsilon_t
$$

Because I'm working with time series data, I conducted several tests to ensure unbiased estimates. The daily attendance counts displayed both conditional heteroskedasticity (Breusch-Pagan $\chi^2$ = 889.97, p<0.01) and positive serial correlation (Durbin-Watson = 1.69, p<0.01). To maintain sound standard error measures, I report Newey–West heteroskedasticity and autocorrelation‑consistent (HAC) standard errors with a seven day bandwidth. Additionally, an autocorrelation function (ACF) test revealed strong AR(1) persistence, necessitating the addition of a lagged dependent variable.

# Results
```{r first regression, include=FALSE}

model <- lm(total ~ lag(total) + day + week + month + quarter + day_after_closed + finals_week + dead_week, count_open)

#adding a lag reduced AR(1)
summary(model)

.robust <- sqrt(diag(NeweyWest(model, lag = 7)))

```

```{r day regression stargazer, results = "asis"}
.s1Variables <- c("Constant", "dayTuesday", "dayWednesday", "dayThursday", "dayFriday",  "daySaturday",  "daySunday", "lag\\(total\\)")

cat("\\begin{center}\n") 
stargazer(model,
          type = "text",
          keep = .s1Variables,
          order = .s1Variables,
          se = list(.robust),
          keep.stat = c("n", "adj.rsq", "ser", "f"),
          digits = 3,
          dep.var.labels = "Total Attendance",
          covariate.labels = c("Intercept", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday", "Total Students (t-1)"),
          title = "Intraweek Effects",
          header = FALSE,
          no.space = TRUE,
          float = FALSE)
cat("\\end{center}\n")
```

Controlling for AR(1) persistence, the model predicts a statistically significant monotonic decline as the week progresses. The intercept, representing a Monday in fall term, not during finals/dead week, and not after a closure, expects 2,779 students on average, that declines to an intraweek low on Saturday. The lagged attendance variable predicts that 75% of Monday's "surprise", if any, carries on to the next day. For example, if there are 1,000 more patrons on Monday, we expect ~750 more than expected on Tuesday. These outcomes reinforce the idea that gym patrons are not acting as rational agents when deciding how to spend their time. With a consistently decreasing attendance as the week progresses, gym attendees are going when their time is the most valuable (the beginning of the week), and going less as their time becomes less valuable. 


```{r quarter regression stargazer, results = "asis"}
.s2Variables <- c("Constant", "week", "dead_week", "finals_week", "quarterBreak", "quarterWinter", "quarterSpring", "quarterSummer")

cat("\\begin{center}\n") 
stargazer(model,
          type = "latex",
          keep = .s2Variables,
          order = .s2Variables,
          se = list(.robust),
          keep.stat = c("n", "adj.rsq", "ser", "f"),
          digits = 3,
          dep.var.labels = "Total Attendance",
          covariate.labels = c("Intercept", "Week of the Term", "Finals Week", "Week Before Finals", "Break", "Winter", "Spring", "Summer"),
          title = "Long Term Effects",
          header = FALSE,
          no.space = TRUE,
          float = FALSE)
cat("\\end{center}\n")
```

Longer term effects showed more inconsistent significance. The week of the term coefficient reports a significant but negligible negative relationship with attendance, while the finals week variable predictably reports lower attendance. 




The F-statistic reports a statistically significant value of 780, rejecting the null hypothesis that all coefficients are zero.  Coupled with an adjusted $R^2$ of 0.872, this model accounts for a substantial share of day to day variation in gym attendance.

# Conclusion



```{r Attendance Plot, fig.align='center', fig.height=3, fig.width=5, include=FALSE}
.plot2024 <- count %>%
  mutate(year = year(date)) %>%
  filter(year == 2024) %>%
  mutate(week = floor_date(date, "week", week_start = 1)) %>%
  group_by(week) %>% 
  summarise(avg_total = mean(total, na.rm = TRUE), .groups = "drop") %>% 
  ggplot(aes(week, avg_total)) +
  geom_line() +
    labs(x = NULL,
         y = "Average Daily Visits",
         title = "Weekly Average Gym Attendance – 2024") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    plot.subtitle = element_text(size = 7),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 7)
  )

.plot2024
```
