---
title: "thesis"
author: "Hunter Wright"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Topics to cover:
- loss aversion
- herd behavior
- rank members by how valuable their time is

Questions: 

For many people, going to the gym is hard, and more importantly, it takes time. Is there a significant effect of certain events like finals on gym attendance? 

What trends in attendance do we observe across different time periods? 

How do different membership types behave differently in response to certain conditions?



Dummy variables to add:

finals season (dead week + finals week) *
weekend
weekday
vacation *
summer quarter
SRC maintenance closure (september)
COVID-19 Closure
weather (rainy or not) *
football gamedays *
hayward prefontaine classic *
track and field championships/eugene marathon *


```{r read csv}

count <- read.csv("cleanData/count.csv")
members <- read.csv("cleanData/membership.csv")

```
```{r rename count columns}

count <- count %>%
  rename(am5 = "X5.00AM", am6 = "X6.00AM", am7 = "X7.00AM", am8 = "X8.00AM", am9 = "X9.00AM", am10 = "X10.00AM", am11 = "X11.00AM", pm12 = "X12.00PM", pm1 = "X1.00PM", pm2 = "X2.00PM", pm3 = "X3.00PM", pm4 = "X4.00PM", pm5 = "X5.00PM", pm6 = "X6.00PM", pm7 = "X7.00PM", pm8 = "X8.00PM", pm9 = "X9.00PM", pm10 = "X10.00PM", pm11 = "X11.00PM")
```



```{r add dummies}

members <- members %>%
  mutate(day = weekdays(as.Date(date))) %>%
  relocate( day, .after = date)
count <- count %>%
  rename(date = Date) %>%
  mutate(day = weekdays(as.Date(date))) %>%
  relocate( day, .after = date)

```


