---
title: "thesis"
author: "Hunter Wright"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(pacman)
p_load(tidyverse)
```

Topics to cover:
- loss aversion
- herd behavior
- rank members by how valuable their time is

Questions: 

For many people, going to the gym is hard, and more importantly, it takes time. Is there a significant effect of certain events like finals on gym attendance? 

What trends in attendance do we observe across different time periods? 

How do different membership types behave differently in response to certain conditions?



Dummy variables to add:

finals season (dead week + finals week) *
vacation *
summer quarter
SRC maintenance closure (september)
weather (rainy or not) *
football gamedays *
hayward prefontaine classic *
track and field championships/eugene marathon *


```{r read csv}

count <- read.csv("cleanData/count.csv")
members <- read.csv("cleanData/membership.csv")

```
```{r rename count columns}

count <- count %>%
  rename(am5 = "X5.00AM", am6 = "X6.00AM", am7 = "X7.00AM", am8 = "X8.00AM", am9 = "X9.00AM", am10 = "X10.00AM", am11 = "X11.00AM", pm12 = "X12.00PM", pm1 = "X1.00PM", pm2 = "X2.00PM", pm3 = "X3.00PM", pm4 = "X4.00PM", pm5 = "X5.00PM", pm6 = "X6.00PM", pm7 = "X7.00PM", pm8 = "X8.00PM", pm9 = "X9.00PM", pm10 = "X10.00PM", pm11 = "X11.00PM")
```



```{r add dummies}

# create weekday column
members <- members %>%
  mutate(day = weekdays(as.Date(date))) %>%
  relocate( day, .after = date) 
count <- count %>%
  rename(date = Date) %>%
  mutate(day = weekdays(as.Date(date))) %>%
  relocate( day, .after = date)

# weekday/weekend dummy
members <- members %>%
  mutate(weekend = ifelse(day == "Saturday" | day == "Sunday", 1, 0))
count <- count %>%
  mutate(weekend = ifelse(day == "Saturday" | day == "Sunday", 1, 0))

# covid dummy
members <- members %>%
  mutate(covid = ifelse("2020-03-18" <= date & date < "2020-09-21", 1, 0))
count <- count %>%
  mutate(covid = ifelse("2020-03-18" <= date & date < "2020-09-21", 1, 0))

#closed dummy (total entrances = 0)
members <- members %>%
  mutate(closed = ifelse(students == 0, 1, 0)) #no students: probably closed
count <- count %>%
  mutate(closed = ifelse(total < 10, 1, 0)) # account for closures with pe_staff present

#closure numbers within 20 observations of each other
members %>%
  filter(students == 0) %>%
  nrow() #655

count %>%
  filter(total < 10) %>%
  nrow() #672



```

```{r}

```








