---
title: 'University of Oregon Gym Attendance: An Analysis of Student Behavior and Decision
  Making'
author: "Hunter Wright"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library(pacman)
p_load(tidyverse, forcats, kableExtra, stargazer, tseries, lmtest, sandwich, scales, broom)

count <- read.csv("cleanData/count.csv")
members <- read.csv("cleanData/membership.csv")

```
```{r factor releveling, include=FALSE}
members <- members %>%
  mutate(day = fct_relevel(day, "Monday")) %>%
  mutate(month = fct_relevel(month, "October")) %>%
  mutate(quarter = fct_relevel(quarter, "Fall")) %>%
  mutate(all_students = students + student_offterm) %>%
  relocate(all_students, .after = student_offterm)

members$date <- as.Date(members$date)

members_open <- members %>%
  mutate(closed = ifelse(students < 10, 1, 0)) %>%
  filter(closed == 0) %>%
  filter(quarter != "Break") %>%
  rename(term_week = week)
```

# Introduction
  The health and fitness industry has seen substantial growth since COVID-19 pandemic restrictions were lifted. According to the 2024 Health and Fitness Association Global Report, key developed markets around the world are seeing record high rates of gym memberships among their adult population. The United States leads with 23.7% of surveyed adults saying they belonged to a gym or health club. On top of this, markets with historically lower participation are seeing record growth as health and fitness enters the forefront of the public consciousness. According to an article by Fortune, young adults are going at an even higher rate, and are responsible for pushing attendance to nearly double pre-pandemic levels. 38% of survey respondents ages 13-28 (Generation Z) said they were a member at a gym or health club. A majority of undergraduate and graduate college students fall into this age range, and most living close to or on campus find themselves near a college gym. Despite increased gym accessibility, college students have a wide variety of responsibilities and commitments soaking up large amounts of time during the week that make consistent attendance difficult. Consequently, though gym attendance rates are rising worldwide, going remains a privilege to those that can afford to spend time there. This makes college students a natural population to study: young and increasingly health-conscious but facing intense time constraints and competing priorities.
  
  Attending a university is a pivotal point in an individual’s life that introduces new friends, responsibilities, and experiences. For many, it is their first time living independently and making decisions on their own. Naturally, an individual's years at university are a critical habit-formation window where they adopt habits and behaviors that they’ll maintain for the rest of their lives. Regular physical activity is one such behavior with both short and long-term benefits that can drastically improve one’s quality of life. Yet even with access to on-campus gyms and increased awareness of fitness, students often struggle to follow through with fitness plans and maintain consistent exercise regimens.
  
  Every day, students are presented with decisions with ramifications that affect their future. Studying, working, relaxing, and other responsibilities all take time out of everyone’s day. These responsibilities all compete with each other for spots in college students’ extremely limited time on a daily basis. Combine this with a lack of motivation, lack of experience, or any other issue that might make it difficult for a student to go to the gym, and it becomes apparent that many students may fail to follow through with consistent exercise plans, ultimately affecting their long term health.
  
  According to the University of Oregon's Student Recreation Center (SRC) website, “The Student Recreation Center is consistently busy Monday–Thursday during the term… Friday–Sunday, the Rec has much less traffic.” Most students have busy schedules during the week, and much more time on the weekend, so why does SRC traffic decrease when students have more time to go?  In this thesis, I use daily patron entrance data gathered by the SRC to analyze short and long-term trends in attendance and discuss how students struggle to stick with fitness plans and display behavior consistent with present bias when choosing whether or not to go to the gym. 
  
  After regressing gym attendance on variables like day of the week, season, and term, controlling for variables like previous day's attendance and membership type, I find strong, statistically significant drops in attendance in short term (week by week) time periods. Across longer time periods, most attendance fluctuations are negligible or insignificant. These results reinforce the ideas that students at the University of Oregon struggle to maintain consistent exercise regimens and exhibit behavioral patterns consistent with present bias.
  
The rest of this thesis is organized into the following sections. Section 2 provides a brief literature review consisting of important papers in behavioral economics and related papers studying gym attendance. Section 3 discusses standard model economics, habit formation, and present bias. Section 4 describes the SRC and provides insights into the data used for my analysis. Section 5 and 6 describe the methods of the analysis and its results, respectively. Finally, the last section contains a brief conclusion discussing my findings.


# Literature Review

The primary literature relating to this thesis is Prospect Theory: An Analysis of Decision Under Risk, by Kahneman and Tversky (1979). This paper serves as a critique of expected utility theory by providing evidence that humans do not act completely rationally when presented with decisions that involve risk. While Kahneman and Tversky were focused mainly on monetary decisions under risk, I focus on the scarcity of time as another form of currency, where students risk the development of their human capital based on how they utilize limited time. The concept of mental accounting was first developed by Richard Thaler in his 1985 paper "Mental Accounting and Consumer Choice." Thaler introduces a theory that explains how individuals make irrational decisions when spending money because they don't treat money as fungible. Present bias was formally introduced in "Doing It Now or Later" by O'Donoghue and Rabin (1999).  The authors described "the human tendency to grab immediate rewards and to avoid immediate costs in a way that our 'long-run selves' do not appreciate." They called this tendency **present-biased preferences**, and partially attributed it to impatience. The concept of loss aversion was first introduced in Kahneman and Tversky's "Prospect Theory," but further refined in "Loss Aversion in Riskless Choice: A Reference-Dependent Model" by Tversky et. al (1991). Other similar literature related to gym attendance focus mainly on individuals' tendency to overestimate their future attendance. DellaVigna and Malmendier (2006) found that people who bought monthly memberships tended to go infrequently and indirectly pay more per visit than if they had bought a 10 day pass. They also found that monthly membership customers were more likely than yearly members to continue their membership for longer than a year, despite paying premiums for the option to cancel month by month. Acland and Levy (2015) reported that subjects in their experiment exhibited "evidence of partial naiveté with respect to present bias" when making projections about their future gym attendance. The subjects similarly overestimated their future attendance. 

# Concepts in Behavioral Economics Theory

Behavioral economics is the study of how people in the real world approach decisions in an irrational manner. Contrary to what standard model economics believes, people don't disregard emotion, habit, and other feelings when making decisions. Though behavioral economics primarily focuses on how individuals make decisions with limited money, this thesis shows that individuals may treat limited time in a similar manner.  

### The Standard Economic Model

The standard economic model predicts that all individuals are rational agents that maximize their utility subject to a constraint. Because I aim to analyze how students utilize their time, in this thesis, the agent's constraint is the number of hours in a day. If this agent was behaving rationally, they would plan and schedule times to go to the gym every week and stick to that plan throughout the year. If students at the SRC behaved this way, we would expect gym attendance to be consistent across all time periods, with a possible slight uptick on weekends when students have more time to go. However, as Figure 1 shows, this tends not to be the case.

```{r average fall 2024 attendance plot, fig.height=3, fig.width=5, fig.align = 'center'}

.att_plot <- members %>%
  mutate(year = year(date)) %>%
  filter(year == 2024,
         quarter == "Fall") %>%
  group_by(day) %>%
  summarize(average = mean(all_students)) %>%
  ungroup() %>%
  mutate(
    day = factor(day, levels = c("Monday","Tuesday","Wednesday",
                                "Thursday","Friday","Saturday","Sunday"))) %>%
  ggplot(aes(x = day, y = average, group = 1)) + 
  geom_line() +
  geom_point(size = 2) +
  scale_y_continuous(labels = comma_format()) +
  labs(
    x = NULL,
    y = "Attendance",
    title = "Figure 1: Average SRC Daily Student Attendance, Fall 2024"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 7),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 7)
  )

.att_plot

```

### Present Bias
 
Individuals experiencing a dilemma with ramifications that unfold over time (known as intertemporal choice) may engage in temporal discounting, where they place disproportionately less value on future rewards, even if they are greater than present rewards. This is known as *present bias*. For example, someone may choose to use their disposable income to consume goods (like buying a car), instead of putting it in a savings account, even though investing will ultimately net them more utility in the long run. In this thesis, I aim to interpret present bias in how individuals utilize their time. For many people, going to the gym is hard and its payoffs take time. If someone only has two days a week to relax, they might skip the gym on the weekend and spend their time relaxing, because the immediate gratification of relaxing is worth more to them than the future benefit of a consistent gym routine. This could explain why average student attendance is so much lower on weekends, and is a significant departure from the standard economic model's prediction, which interprets this behavior as irrational.

# Data

### Setting

The University of Oregon is the state of Oregon's flagship public university. As a Division I school featuring more than 30 club and intramural sports, fitness is a large part of many student's lives. Whether they're rooting for the Ducks, participating in organized sports, or simply exercising on campus, students at UO are incredibly active.
The University of Oregon’s Student Recreation Center is the primary facility on campus for students to engage in physical activities like weightlifting, swimming, basketball, and other indoor sports. Students, faculty, staff, and community members alike enjoy more than 36,000 square feet of strength training space, a 12-lane swimming pool, and nearly two dozen courts and gymnasiums, among many other amenities. During normal hours, the SRC is open from 6am - 11pm on weekdays and 9am-9pm on weekends. 

### Data Collection

Almost every person that enters the SRC must scan their ID through one of the turnstiles. These machines record the time and date of every person who walks through. Table 1 shows the summary statistics for daily entries between January 2015 and December 2024 for days where the SRC is open. This accounts for about 82% of the days in the data, which includes a period between March 18th, 2020 and September 17th, 2020 where the SRC closed due to COVID-19. A day was marked as 'closed' if no students entered the gym on a particular day.

```{r all students summary statistics table}

# closed days are removed
.summary_stats <- members_open %>%
  summarise(
    Mean = mean(all_students),
    Median = median(all_students),
    SD = sd(all_students),
    Min = min(all_students),
    Max = max(all_students)
  )

kable(.summary_stats, caption = "Student Attendance Summary Statistics") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(0, align = "c") %>%
  row_spec(1, align = "c")
```

Table 2 shows the average daily proportion of members. Students make up the vast majority of the SRC's daily entrants. It should be noted that "Off-Term Students" are simply students who are not regarded as full-time students because they are taking less than 12 credits of classes. This number is skewed by the heavy increase in this membership type that occurs every summer when students continue to use the SRC while on summer break.

```{r membership summary statistics table}


.member_columns <- c("students", "student_offterm", "faculty_staff", "pe_staff", "alumni", "nonmembers", "community_members", "other")

.mem_sumstats <- members_open %>% 
  summarise(
    across(all_of(.member_columns),
           ~ mean(.x / total, na.rm = TRUE)) 
  ) %>%
  rename("Student" = students,
         "Faculty/Staff" = faculty_staff,
         "PE Staff" = pe_staff,
         "Alumni" = alumni,
         "Nonmember" = nonmembers,
         "Community Member" = community_members,
         "Off-Term Student" = student_offterm,
         "Other" = other) %>% 
  pivot_longer(everything(),
               names_to  = "Membership Type",
               values_to = "Average Daily Share") %>% 
  mutate(`Average Daily Share` = percent(`Average Daily Share`, accuracy = 0.1)) %>%
  arrange(desc(`Average Daily Share`))

.mem_sumstats$`Membership Type` <- factor(.mem_sumstats$`Membership Type`, 
    levels = c("Student", "Off-Term Student", "Faculty/Staff", "Community Member", 
               "Alumni", "Nonmember", "PE Staff", "Other"))

.mem_sumstats <- .mem_sumstats %>% arrange(`Membership Type`)


# need to fix row separations
kable(.mem_sumstats, caption = "Average Daily Entrances by Member") %>%
  kable_styling(full_width = FALSE, latex_options = c("hold_position")) %>%
  row_spec(0:8, align = "c")
```

Two data sets were collected and used for this analysis. The first displayed simple daily totals grouped by time of day, while the other displayed these totals grouped by membership type. There is a discrepancy in daily total entrances between the data sets, where the membership data set tends to under count daily entrances, averaging about 277 fewer entrances a day. The reason for this is unknown, but it can be reasonably assumed that this discrepancy has a negligible effect on attendance trends and behavior patterns. The membership data is the only data used in this analysis, as student behavior is the only behavior being analyzed. Observations where the SRC is closed for any reason are removed from the analysis.

# Methods

To analyze both short and long‑run attendance dynamics, I regress daily student attendance on lagged attendance, weekday, week‑of‑term, month, quarter, post‑closure status, and finals/dead‑week indicators. 

$$
\hat{\text{attendance}}_{t} = \beta_0 + \beta_1\,\text{attendance}_{t-1} \;+\; \sum_{d=1}^{6}\gamma_d\,\text{Day}_{d,t} \;+\;\beta_2\,\text{Week}_t
\;+\; \sum_{q=2}^{4}\theta_q\,\text{Quarter}_{q,t}
$$ 

$$
\;+\; \beta_3\,\text{DayAfterClosed}_t\;+\; \beta_4\,\text{DeadWeek}_t \;+\; \beta_5\,\text{FinalsWeek}_t \;+\; \varepsilon_t
$$
  The first term in the regression is $\beta_0$, the intercept term. This represents the model’s expected attendance when all other coefficients are equal to zero. $\beta_1$ is the lag term, where t is a subscript for the calendar-day. The coefficient here is a decimal representing what proportion of the previous day's “surprise” or deviation from the expected value is present in today's count. The next term represents day-of-the-week fixed effects: Each day's difference in expected attendance relative to Monday. $\beta_2$ describes the change in students as the quarter progresses. The coefficient represents the change in students on a weekly basis. Next, quarter fixed effects describe the average difference in attendance across different quarters, or terms. There are four quarters, Summer, Fall, Winter, and Spring, and a catch-all, Break, which is active during days that are not a part of any quarter. Fall is the baseline quarter that the other four are compared to. $\beta_3$, $\beta_4$, and $\beta_5$, are all binary variables describing if the SRC was closed the previous day, if it’s dead week (the week before finals week), or finals week respectively. All are equal to one if the requirement is met, and zero if not.
  Because I'm working with time series data, I conducted several tests to ensure unbiased estimates. The daily attendance counts displayed both conditional heteroskedasticity (Breusch-Pagan $\chi^2$ = 186.74, p<0.01) and positive serial correlation (Durbin-Watson = 1.69, p<0.01). To maintain sound standard error measures, I report Newey–West heteroskedasticity and autocorrelation‑consistent (HAC) standard errors with a seven day bandwidth. Additionally, an autocorrelation function (ACF) test revealed strong AR(1) persistence, necessitating the addition of a lagged dependent variable.

# Results
```{r first regression}
options(scipen = 2)

# level regressions
.model0  <- lm(all_students ~ day + term_week + quarter + day_after_closed + finals_week + dead_week, members_open)

bptest(.model0)
dwtest(.model0, alt="less")


.model1 <- lm(all_students ~ lag(all_students) + day + term_week + quarter + day_after_closed + finals_week + dead_week, members_open)

# level robust SEs
.robust0 <- sqrt(diag(NeweyWest(.model0, lag = 7)))
.robust1 <- sqrt(diag(NeweyWest(.model1, lag = 7)))

members_open <- members_open %>% 
  mutate(log_students = log(all_students),
         lag_log_students = lag(log_students))

.model0_log <- lm(log_students ~ day + term_week + quarter + day_after_closed + finals_week + dead_week, data = members_open)

.model1_log <- lm(log_students ~ lag_log_students + day + term_week + quarter + day_after_closed + finals_week + dead_week, data = members_open)

.robust0_log <- sqrt(diag(NeweyWest(.model0_log, lag = 7)))
.robust1_log <- sqrt(diag(NeweyWest(.model1_log, lag = 7)))

.s1Variables <- c("Constant", "dayTuesday", "dayWednesday", "dayThursday", "dayFriday",  "daySaturday",  "daySunday", "^lag")

.s2Variables <- c("Constant", "term_week", "quarterWinter", "quarterSpring", "quarterSummer", "dead_week", "finals_week")
```
Both tables represent two sets of predictors from the same model. The first one reports day-of-the-week fixed effects with and without a one day lagged variable, representing short term trends in attendance. The second table reports week of the term and quarter fixed effects, the dead week, and finals week dummies, with and without a one day lagged variable. This represents attendance trends over a longer term, as each quarter of the academic year at the University of Oregon is 11 weeks, or 12 in the summer.

```{r day by day regression stargazer, results = "asis"}
cat("\\begin{center}\n")
cat("\\text{Table 3: Short Term Trends}\\\\\n")
stargazer(.model0,        
          .model1,
          .model0_log,
          .model1_log,
          type = "latex",
          keep = .s1Variables,
          order = .s1Variables,
          se = list(.robust0, .robust1, .robust0_log, .robust1_log),
          keep.stat = c("n", "adj.rsq"),
          digits = 3,
          dep.var.labels = c("Student Attendance", "Log Student Attendance"),
          column.labels = c("No Lag",
                            "With Lag",
                            "No Lag",
                            "With Lag"),
          covariate.labels = c("Intercept", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday", "Lagged Student Attendance (t-1)", "Lagged log-Attendance (t-1)"),
          title = "Intraweek Effects",
          header = FALSE,
          no.space = FALSE,
          float = FALSE,
          font.size = "footnotesize")
cat("\\end{center}\n")
```

The lagless model shows a strong statistically significant negative drop in attendance as the week progresses, culminating in a more than 2,000 student difference in attendance between Monday and Saturday. Using the lag to account for strong AR(1) persistence, the second model finds even steeper results. The intercept, representing a Monday in fall term, not during finals/dead week, and not after a closure, expects 2,404 students on average, that declines to an intraweek low on Saturday. The lagged attendance variable predicts that about 72% of the previous day's "surprise", if any, carries on to the next day. For example, if there are 1,000 more entrants on Monday than usual, we expect ~724 more than usual on Tuesday. Controlling for this "habit" raises Adj. $R^2$ from 0.7 to 0.87, indicating that yesterday's turnout is a strong predictor for today's but isn't able to carry the prediction alone.

```{r bar chart, fig.height=3, fig.width=5, fig.align = 'center'}
.bar <- tidy(.model1) %>% 
  filter(term %in% c("dayTuesday","dayWednesday","dayThursday",
                     "dayFriday","daySaturday","daySunday")) %>% 
  mutate(term  = str_remove(term, "^day"), 
         term  = fct_relevel(term, 
                             "Tuesday","Wednesday","Thursday",
                             "Friday","Saturday","Sunday"))

ggplot(.bar, aes(term, estimate)) +
  geom_col(fill = "black", width = .4) + 
  labs(x = NULL,
       y = "Student Attendance Estimates") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 10)) +
  scale_x_discrete(labels = c(
    "Tuesday" = "Tue",
    "Wednesday" = "Wed",
    "Thursday" = "Thu",
    "Friday" = "Fri",
    "Saturday" = "Sat",
    "Sunday" = "Sun"
  ))
```

These coefficients reinforce the idea that college students are not acting as rational agents when deciding how to spend their time. With a consistently decreasing attendance as the week progresses, students are going when their time is the most valuable (the beginning of the week), and going less as their time becomes less valuable (the weekend).

```{r quarter regression stargazer, results = "asis"}
cat("\\begin{center}\n")
cat("\\text{Table 4: Long Term Trends}\\\\\n")
stargazer(.model0, .model1, .model0_log, .model1_log,
          type = "latex",
          order = .s2Variables,
          keep = .s2Variables,
          se = list(.robust0, .robust1, .robust0_log, .robust1_log),
          keep.stat = c("n", "adj.rsq"),
          digits = 3,
          dep.var.labels = c("Student Attendance", "Log Student Attendance"),
          column.labels = c("No Lag", "With Lag", "No Lag", "With Lag"),
          covariate.labels = c("Intercept", "Week of the Term", "Winter", "Spring", "Summer", "Week Before Finals", "Finals Week"),
          title = "Long Term Trends",
          header = FALSE,
          no.space = FALSE,
          float = FALSE)
cat("\\end{center}\n")
```

Long term trends, while still mostly significant, are much weaker in comparison to day by day changes. Controlling for AR(1) persistence, there are sizable coefficients on break, spring, and finals week variables. Break understandably sees a sizable drop in attendance as many students go home. Spring quarter in Eugene is notorious for students missing classes, going outside, and enjoying the sun, which explains the negative relationship. Understandably, the finals week estimate is -572, which correlates with students schedules being dominated by studying and preparation for finals. The opportunity cost of all hours during finals week is much higher, leading to lower attendance on average. Weakly significant but negligible estimates constitute the rest of the table.


# Conclusion





# Citations

Acland, D., & Levy, M. R. (2015). Naiveté, projection bias, and habit formation in gym attendance. Management Science, 61(1), 146–160.

DellaVigna, Stefano and Ulrike Malmendier. (2006). "Paying Not to Go to the Gym." American Economic Review, 96 (3): 694–719.

Kahneman, D., & Tversky, A. (1979). Prospect theory: An analysis of decision under risk. Econometrica, 47(2), 263-291.

O'Donoghue, Ted, and Matthew Rabin. (1999). "Doing It Now or Later." American Economic Review 89 (1): 103–124.

Thaler, Richard. (1985). Mental Accounting and Consumer Choice. Marketing Science, 4(3), 199–214.

Tversky, Amos, & Daniel Kahneman. (1991). Loss Aversion in Riskless Choice: A Reference-Dependent Model. The Quarterly Journal of Economics, 106(4), 1039–1061.

“The 2024 HFA Global Report.” Health & Fitness Association, https://www.healthandfitness.org/publications/the-2024-hfa-global-report/.

Bradley, Jordyn. “Is Your Gym Overflowing? A New Study Shows Gym Use Is Nearly Double Pre-Pandemic Levels and Gen Z Is Driving the Trend.” Fortune Well, https://fortune.com/well/article/gym-use-increase-pandemic-gen-z/.

https://admissions.uoregon.edu/uo-facts


